Deciding Bitvector Termination with Program Synthesis

### 2014-06-25 papers/termination/paper.tex
First version of autogenerated results table
preliminaries

Limitations: No recursion.

@@

\section{Motivational Examples} \label{sec:motivation}
In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
Now, we provide some simple examples that break those assumptions and cannot therefore be handled by state of the art tools.

@@

Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
A program not falling in this category is given in Figure~\ref{fig:motivation.a}.
%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
In order to find a ranking function for this example, it is necessary to take into account
the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers.
Our technique finds that a possible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but it can not decrease indefinitely as it is bounded from below.
This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.

The program in Figure~\ref{fig:motivation.b} breaks a different assumption made by the majority of termination analyses, namely that \emph{ranking functions are linear}.
Our technique proves that this loop terminates by finding that $|x|$ is a possible ranking function for it.

Figure~\ref{fig:motivation.c} invalidates the assumption that \emph{integer semantics and bit-vector semantics are equivalent}. 
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

The same assumption is broken by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: in the first phase $y$ increases
until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
This example is taken from \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
However, when considering the bit-vector semantics, this program does not terminate as..

The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break the assumption that \emph{real semantics and float semantics are equivalent}. 
% ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
This assumption may lead to erroneous diagnosis of a program's terminating behaviour.% as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
While the former does not terminate for reals, but does for floats, the latter 
terminates for reals, but does not for floats.

@@

\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
\todo{linear program}

@@

@@

If $Y = Z^m$, we say that the ranking function is \emph{lexicographic},
and require that the total order imposed on $Y$ is the lexicographic ordering
induced on tuples of $Z$'s.  We note that some termination arguments
require lexicographic ranking functions, or equivalently, ranking functions
whose co-domain is the ordinals, rather than just $\mathbb{N}$.

>>>>

Limitations: 
\begin{itemize}
\item No recursion.
\item Nested loops.
\end{itemize}

@@

\section{Motivational Examples} \label{sec:motivation}
Figure~\ref{fig:handletable} illustrates the most common assumptions made by existent termination analyses:
\begin{itemize}
\item[(i)] programs are linear.
\item[(ii)] terminating programs have linear termination arguments.
\item[(iii)] bit-vectors semantics and mathematical integers semantics are equivalent.
\item[(iv)] floats semantics and mathematical reals semantics are equivalent.
\end{itemize}

%In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
Next, we show how these assumptions are broken by even simple programs that can't therefore be handled accurately by state of the art termination tools.

@@

%Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
The loop in Figure~\ref{fig:motivation.a} breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
%In order to find a ranking function for this example, it is necessary to take into account
%the semantics of the bit-wise AND operator.
Our technique finds that a possible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but cannot decrease indefinitely as it is bounded from below.
This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.

Assumption (ii) is broken by the loop in Figure~\ref{fig:motivation.b}.
We prove that this loop terminates by finding that the non-linear function $R(x) = |x|$ is a possible ranking function.

Figure~\ref{fig:motivation.c} invalidates assumption (iii). 
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

Assumption (iii) is broken again by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: 
in the first phase $y$ increases until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
This example is taken from \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
However, when considering the bit-vector semantics, this program does not terminate as..\todo{change this example}

The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break assumption (iv) as their terminating behaviour is different 
for floats and mathematical reals. 
%This assumption may lead to erroneous diagnosis of a program's terminating behaviour.
The former does not terminate for floats (if $x$ is sufficiently large, the rounding makes the subtraction negligible), but does for reals. Conversely, the latter 
doesn't terminate for reals, but does so for floats (due to rounding errors $x$ eventually becomes sufficiently small 
such that the closest representable number is 0.0). \todo{can we prove non-termination?}

@@

\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
We represent programs as transition systems with state space $X$, transition relation $T \subseteq X \times X$
and the set of variables $$

@@

\begin{definition}[Linear ranking function]
A linear ranking function $R: X \to Y$ 
with $\dim(X) = n$ and $\dim(Y) = m$ is of the form:
is one that can be expressed as an $n \times m$ matrix $M$:
$$f(\vec{x}) = M\vec{x}$$
\end{definition}

\begin{definition}[Lexicographic ranking function]
If $Y = Z^m$, we say that a ranking function $R: X \to Y$ is lexicographic
and maps each state in $X$ to a tuple of values such that the loop transition leads to a decrease with
respect to the lexicographic ordering for this tuple.
The total order imposed on $Y$ is the lexicographic ordering
induced on tuples of $Z$'s. 
\end{definition}

For lexicographic ranking functions, no function may increase unless a function of a lower index decreases.
Additionally, at every step, there must be at least one function that decreases.
We note that some termination arguments require lexicographic ranking functions, or equivalently, ranking functions
whose co-domain is the ordinals, rather than just $\mathbb{N}$.

In the rest of the paper, we consider any ranking function $R: X \to Z^m$ to be lexicographic, and the situation when $m=1$ to be a special case.

@@
